---
title: "Applying Lasso/Ridge Regression Upon NFL Data"
subtitle: "Predicting Team Wins Based On QB Play"
author: "Jonathan Sears, Jasmine Cole, Noah Miller"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"}

## Introduction

Under the stadium lights, the smell of popcorn and hot dogs and the band
playing the “Fight Song” as one final play runs the last 3 seconds of
the 4th quarter, that is the aesthetic of an American Football game.
Fans come together to support, cheer and hope for their team's victory.
But what if fans or people in general had the ability to go into a match
where they could predict the outcome? Being able to predict outcomes in
the National Football League (NFL) through multi-dimensional data
requires many different features and data such as home field advantage,
player performance, team statistics, and game conditions which can cause
multicollinearity, being a negative impact on the least square estimator
in regression due to correlation between variables
[@mcdonald2009ridge]. 

With the implementation of a machine learning regression model such as
LASSO (Least Absolute Shrinkage and Selection Operator), and Ridge we
can simplify the variables and focus only on the important aspects that
help to accurately predict wins. Both methods solve a modified
optimization problem of the form:

$$\hat{\boldsymbol{\beta}}
=
\arg\min_{\boldsymbol{\beta}}
\left\{
\|\mathbf{y}-\mathbf{X}\boldsymbol{\beta}\|_2^2
+
\lambda P(\boldsymbol{\beta})
\right\}$$

where $\hat{\boldsymbol{\beta}}$ represents the estimated coefficient
vector, $\arg\min_{\boldsymbol{\beta}}$ represents the values of
$\boldsymbol{\beta}$ that minimize the objective function, $\mathbf{y}$
represents the observed response vector, $\mathbf{X}$ represents the
design matrix of predictor variables, $\boldsymbol{\beta}$ represents
the regression coefficient vector,
$\|\mathbf{y}-\mathbf{X}\boldsymbol{\beta}\|_2^2$ represents the
residual sum of squares, $\lambda$ represents the regularization
parameter controlling the strength of the penalty, and
$P(\boldsymbol{\beta})$ represents the penalty function applied to the
coefficients.

The difference between LASSO and Ridge regression comes from how their
penalties are calculated mathematically. LASSO uses a shrinkage
technique called L1 regularization, where it zeros out the least
important variables, leaving only variables that are optimal for
accurate prediction. Ridge regression uses a shrinkage technique called
L2 regularization where it helps reduce multicollinearity without
removing variables from the original set of independent variables. This
helps with some of the distortion of the predictors. Another area where
ridge differs from LASSO is that in the end and final model, ridge
retains all variables . LASSO is a “step-up” from Ridge allowing for key
variable selection, where Ridge can struggle with its limitations due to
retention of all variables. This highlights that LASSO’s ability to
produce a subset of “key-variables” from a high-dimensional dataset,
makes it highly efficient.

Previous research examining LASSO regression has consistently
demonstrated its usefulness in datasets containing large numbers of
predictors and correlated variables. Across applications in credit risks
awareness [@geraldo2022credit], genetics, and other predictive modeling
fields, LASSO has been shown to improve model interpretability by
shrinking insignificant coefficients toward zero while retaining the
most influential predictors [@bhattacharyya2025lasso; @jana2021football;
@lockhart2014lasso]. Sports outcomes in LASSO include ability to
identify home-field advantage as a key-variable by narrowing down
recruitment and box-score metrics [@south2020forecasting]. Similarly,
LASSO was utilized to significantly reduce computational complexity that
identified an optimal configuration of motion sensors
[@laufer2023sensor]. Such reduces model complexity and can improve
predictive performance compared to traditional regression methods
[@alshqaq2023diagnostic; @bhattacharyya2025lasso; @lockhart2014lasso].
Studies have also shown that regularization techniques such as Ridge
regression can effectively address multicollinearity by reducing
coefficient variance without removing predictors from the model. One
limitation discussed within the literature is that highly correlated
variables can influence LASSO's variable selection process, potentially
causing relevant predictors to be excluded [@lockhart2014lasso].
Overall, the findings suggest that regularized regression methods
provide a useful framework for prediction problems involving numerous
and potentially correlated variables, making them applicable to NFL team
performance analysis.

It is important to note, however, that LASSO has experienced some
limitations in its performance ability in variable selection
specifically with highly correlated variables and complex non-linear
interactions. For example, LASSO regression was compared against XGBoost
to determine which methodology proved to be the most accurate at
predicting fare prices for online cab services, in which XGBoost
outperformed LASSO because “LASSO failed to make the grouped selection
as it only took one variable from a group and ignored the remaining
variables,” [@tarun2026xgboost]. Despite this specific limitation,
across multiple studies LASSO proves to be consistent in its ability to
minimize prediction error and utilize mathematical frameworks to isolate
key critical variables.

The main goal of this report is to examine the impact of
multicollinearity on predicting NFL team success and to examine whether
regularized regression models are better than a linear regression model.
This will be seen by comparing a standard linear regression, Ridge
regression, and LASSO regression ability to predict the number of
regular-season wins by an NFL team using that team’s quarterback's
performance statistics. Predictor variables include a quarterback’s
season standard statistics, as well as expected points added (EPA), a
metric that measures how many net points a quarterback adds to their
team's scoring potential on passing plays, compared to the
league-average expectation for any given down, distance, and field
position. This report will also aim to find which quarterback
performance statistics are most influential in predicting team
performance. Understanding the impact of a quarterback’s performance on
a team's success and the most influential quarterback performance
statistics provide valuable information for NFL analysts, coaches, and
teams trying to find players that contribute to team success and improve
team performance.

## Methods

Following base cleaning of the data, numerical values were scaled to a
mean of 0 and a standard deviation of 1. The data were standardized
based upon:

$$x_{ij}^{\text{scaled}}
=
\frac{x_{ij}-\mu_j}{\sigma_j}$$

where $x_{ij}$ represents value of feature j for observation i, $\mu_j$
represents mean of feature j, and $\sigma_j$ represents standard
deviation of feature j.

A standard linear regression is used as a baseline, calculated by:

$$y_i = \beta_0 + \sum_{j=1}^{p} \beta_j x_{ij} + \epsilon_i$$

where $y_i$ represents the observed response for observation i,
$\beta_0$ represents the intercept term, $x_{ij}$ represents the value
of predictor variable j for observation i, $\beta_j$ represents the
coefficient associated with predictor, $j$ represents the index of the
predictor variables, and $p$ represents the total number of predictor
variables, and $\epsilon_i$ represents the random error term for
observation i.

For the LASSO and Ridge regression methods, the regularization parameter
($\lambda$) was selected using a 10-fold cross validation, defined as:

$$\mathrm{CV}_{10} = \frac{1}{10}\sum_{k=1}^{10} L_k$$

ensuring a practical 90/10 train-test split, a common approach for
numerical datasets of less than 1000 observations. Within the function,
the dataset is divided into ten equally sized subsets, where nine folds
are used to train the model while the remaining fold is reserved for
validation (testing). The process is repeated until each fold has served
as the validation set exactly once, and the resulting validation errors
are averaged to estimate the model's predictive performance. By
evaluating each candidate value of the regularization parameter
($\lambda$) across all ten folds, what is identified is the value that
minimizes the cross-validated error. The use of 10-fold cross-validation
provides a balance between computational efficiency and reliable model
evaluation, allowing nearly all observations to contribute to both model
training and validation throughout the procedure, without heightened
risk of overfitting or underfitting.

The LASSO regression equation upon the data was calculated by:

$$\hat{\boldsymbol{\beta}}
=
\arg\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(
y_i-\beta_0-\sum_{j=1}^{p}\beta_jx_{ij}
\right)^2
+
\lambda\sum_{j=1}^{p}|\beta_j|
\right\}$$

where $\hat{\boldsymbol{\beta}}$ represents the estimated coefficient
vector, $\lambda$ represents the regularization parameter, and
$\lambda\sum_{j=1}^{p}|\beta_j|$ representing the L1 penalty.

The Ridge regression equation upon the data was calculated by:

$$\hat{\boldsymbol{\beta}}
=
\arg\min_{\boldsymbol{\beta}}
\left\{
\sum_{i=1}^{n}
\left(
y_i-\beta_0-\sum_{j=1}^{p}\beta_jx_{ij}
\right)^2
+
\lambda\sum_{j=1}^{p}\beta_j^2
\right\}$$

where $\lambda\sum_{j=1}^{p}\beta_j^2$ represents the L2 penalty.

To evaluate the predictive performance of the LASSO and Ridge regression
models, RMSE, MAE, and R² were calculated using the cross-validated
predictions. RMSE and MAE quantify the average prediction error, with
RMSE placing greater emphasis on larger errors due to the squared error
term. RMSE is represented by:

$$\mathrm{RMSE}
=
\sqrt{
\frac{1}{n}
\sum_{i=1}^{n}
(y_i-\hat{y}_i)^2
}$$

where n represents the number of observations, $y_i$ represents the
observed response value for observation i, $\hat{y}_i$ represents the
predicted response value for observation i, and $y_i-\hat{y}_i$
represents the prediction error.

MAE provides a measure that is less sensitive to extreme observations,
represented by:

$$\mathrm{MAE}
=
\frac{1}{n}
\sum_{i=1}^{n}
\left|y_i-\hat{y}_i\right|$$

R², however, measures the proportion of variation in the response
variable explained by the model,represented by:

$$R^2 =
1 -
\frac{
\sum_{i=1}^{n}(y_i-\hat{y}_i)^2
}{
\sum_{i=1}^{n}(y_i-\bar{y})^2
}$$

where $\bar{y}$ represents the mean of the observed response values,
$\sum_{i=1}^{n}(y_i-\hat{y}_i)^2$ represents the residual sum of
squares, and $\sum_{i=1}^{n}(y_i-\bar{y})^2$ represents the total sum of
squares.

Lower RMSE and MAE values, coupled with a higher R², indicate a model
that more effectively balances prediction accuracy with the coefficient
shrinkage introduced through regularization.

The R programming language was the primary software used in this
analysis. Data for the analysis was sourced from the nflreadr package; a
repository of NFL data easily accessible within the R environment. For
data preprocessing, the Python programming language package pandas was
used within the R environment, as well as the native dplyr package. The
corrplot, car, and ggplot2 packages were used for the preliminary
analysis and analysis of the final models. Computation and selection of
the LASSO and Ridge-based models was completed through use of glmnet.

## Analysis and Results

### Data Description and Visualization

A study was conducted to determine how quarterback (QB) season
statistics correlates with how many wins a team has in that season. To
determine how QB statistics correlates with team wins in a season, the
dataset was created using publicly available NFL data accessed through
the R package [nflreadr]{.underline}. Weekly quarterback player
statistics and NFL game schedules were collected for the 2015 through
2024 regular seasons.

Quarterback performance data were obtained using the
`load_player_stats()` function. The dataset was filtered to include only
players whose position was quarterback (QB). Weekly statistics were then
aggregated into season-level totals for each quarterback and team. The
aggregated variables included passing yards, passing touchdowns,
interceptions, completions, pass attempts, passing EPA (Expected Points
Added), rushing yards, and rushing touchdowns. In addition, completion
percentage was calculated by dividing total completions by total pass
attempts and multiplying by 100.

Team performance data were collected using the `load_schedules()`
function. Only regular-season games were included by filtering for
`game_type == "REG"`. Wins were calculated separately for home and away
teams by comparing the final scores of each game. Home and away wins
were then combined to compute each team's total number of regular-season
victories for every season from 2015 through 2024.

The team win totals were merged with the quarterback season statistics
using season and team as matching variables. Finally, quarterbacks with
fewer than 200 pass attempts in a season were removed to ensure that the
analysis included only quarterbacks with substantial playing time,
reducing the influence of backup quarterbacks and small sample sizes.

The final dataset contains one observation for each qualifying
quarterback-season and includes both individual performance statistics
and the corresponding team's regular-season win total. The description
of variables are listed below in the chart.

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(corrplot)
library(nflreadr)
library(dplyr)
```

```{r, warning=FALSE, message=FALSE, echo=TRUE}
# Load Data

years = 2015:2024
weekly = load_player_stats(years)
qbs <- weekly %>%
  filter(position == "QB")
qb_season <- qbs %>%
  group_by(player_display_name, season, team) %>%
  summarize(
    passing_yards = sum(passing_yards, na.rm = TRUE),
    passing_tds = sum(passing_tds, na.rm = TRUE),
    interceptions = sum(passing_interceptions, na.rm = TRUE),
    completions = sum(completions, na.rm = TRUE),
    attempts = sum(attempts, na.rm = TRUE),
    passing_epa = sum(passing_epa, na.rm = TRUE),
    rushing_yards = sum(rushing_yards, na.rm = TRUE),
    rushing_tds = sum(rushing_tds, na.rm = TRUE)
  )

qb_season <- qb_season %>%
  mutate(
    completion_pct = (completions / attempts) * 100
  )

write.csv(qb_season, "qb_season_stats.csv", row.names = FALSE)

schedules <- load_schedules(2015:2024)
team_wins <- schedules %>%
  filter(game_type == "REG") %>%
  mutate(
    home_win = ifelse(home_score > away_score, 1, 0),
    away_win = ifelse(away_score > home_score, 1, 0)
  )

home_wins <- team_wins %>%
  group_by(season, team = home_team) %>%
  summarize(wins = sum(home_win, na.rm = TRUE))

away_wins <- team_wins %>%
  group_by(season, team = away_team) %>%
  summarize(wins = sum(away_win, na.rm = TRUE))

team_total_wins <- bind_rows(home_wins, away_wins) %>%
  group_by(season, team) %>%
  summarize(wins = sum(wins))

qb_season <- qb_season %>%
  left_join(team_total_wins,
            by = c("season", "team"))

write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season <- qb_season %>%
  filter(attempts >= 200)


write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season1 <- qb_season %>%
  ungroup() %>%
  select(-player_display_name, -season, -team)
```

```{r}
library(dplyr)
library(knitr)

variable_table <- data.frame(
  Variable = c(
    "passing_yards",
    "passing_tds",
    "interceptions",
    "completions",
    "attempts",
    "passing_epa",
    "rushing_yards",
    "rushing_tds",
    "completion_pct",
    "wins"
  ),
  Description = c(
    "Total number of passing yards recorded by the quarterback during the season.",
    "Total number of passing touchdowns thrown by the quarterback during the season.",
    "Total number of interceptions thrown by the quarterback during the season.",
    "Total number of completed passes by the quarterback during the season.",
    "Total number of pass attempts by the quarterback during the season.",
    "Total Expected Points Added (EPA) generated through passing plays during the season.",
    "Total number of rushing yards gained by the quarterback during the season.",
    "Total number of rushing touchdowns scored by the quarterback during the season.",
    "Completion percentage, calculated as (completions ÷ attempts) × 100.",
    "Total number of regular-season wins recorded by the quarterback's team (response variable)."
  ),
  Type = c(
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Response"
  )
)

kable(variable_table,
      caption = "Table 1: Description of Variables Used in the Analysis")
```

Figure 1: Correlation Plot between Variables

```{r}
correlation_matrix = round(cor(qb_season1, use = "complete.obs"), 2)
corrplot(correlation_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```

According to the correlation plot above, the passing EPA has the
strongest correlation with team wins (r = 0.66). Completion percentage
showed a moderate correlation with wins (r=0.41). The rushing statistics
like rushing touchdowns and rushing yards appeared to have the lowest
positive relationship with wins (average r= 0.22) while interceptions
had almost no linear relationship with wins (r=0.06). 

The analysis also identified substantial multicollinearity among several
passing statistics. Completions and attempts were nearly perfectly
correlated (r = 0.99), while passing yards showed very strong
correlations with completions (r = 0.97) and attempts (r = 0.96). These
findings suggest that some passing variables may provide redundant
information and should be carefully considered during model
development. 

```{r}
library(ggplot2)

ggplot(qb_season1, aes(x = passing_tds, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm",color = "red") + labs(x="Passing touchdowns", y="Team wins", title = "Figure 2: Passing Touchdowns vs Team Wins")
```

```{r}
ggplot(qb_season1, aes(x = passing_yards, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm") + labs(x="Passing yards", y="Team wins", title = "Figure 3: Passing Yards vs Team Wins")
```

Both scatterplots show that passing yards and passing touchdowns have a
relative strong positive correlation to team wins. These scatterplots
show that both variables, passing yards and passing touchdowns have a
positive effect on team wins. The passing touchdowns versus team wins
scatterplot shows a stronger relationship with less variability and
spread in the data comparatively to passing yards versus team wins. The
conclusion can be made that as passing touchdowns increase, team wins
also increase and similarly as passing yards increase, team wins
increase. This conclusion is beneficial in highlightling possible key
variables that strongly influence team wins. \

```{r}
VIF_Visualization_Model <- lm(wins ~ passing_yards + passing_tds + interceptions + completions + attempts + passing_epa + rushing_yards + rushing_tds + completion_pct, data = qb_season1)
library(car)
VIF_Visualization_Values <- vif(VIF_Visualization_Model)
VIF_Visualization_Values
VIF_Visualization_Dataframe <- data.frame(Variable = names(VIF_Visualization_Values), VIF = as.numeric(VIF_Visualization_Values))
library(ggplot2)
ggplot(VIF_Visualization_Dataframe, aes(x = reorder(Variable, VIF), y = VIF)) + geom_col(fill = "steelblue") + geom_hline(yintercept = 5, linetype = "dashed", color = "red") + coord_flip() + labs(title = "Figure 4: Variance Inflation Factors (VIFs) of Dataset Predictors", x = "Predictor", y = "VIF") + theme_minimal()

```

Variance Inflation Factors (VIFs) of the predictors against the response
variable (wins) are modeled based on a simple linear regression created
for the purposes of this visualization. To combat inflated variances due
to predictor multicollinearity, a VIF threshold of 5 was used, exhibited
by the red-dashed line running along the y-axis. Predictors such as
interceptions, rushing_tds, and rushing_yards satisfy this threshold.
Although completion_pct, passing_tds, and passing_epa do not meet the
threshold, they are close enough to at least consider in a final model
if other evidence supports their inclusion. Completions, attempts, and
passing_yards, however, indicate such high variances that they should
likely not be used when refining a model.

### Statistical Modeling

After collecting the quarterback and team data, several preprocessing
steps were performed to prepare the dataset for analysis.

First, weekly quarterback statistics were aggregated into season-level
totals for each quarterback and team from the 2015–2024 NFL regular
seasons. Team regular-season win totals were calculated separately from
the schedule data by summing home and away victories and then merged
with the quarterback statistics using the season and team as matching
variables.

A new variable, completion percentage, was created by dividing total
completions by total pass attempts and multiplying by 100. This provided
an additional measure of passing efficiency beyond the raw completion
and attempt totals.

To ensure that only quarterbacks with meaningful playing time were
included, seasons in which a quarterback attempted fewer than 200 passes
were removed from the dataset. This filtering step reduced the influence
of backup quarterbacks and players with limited appearances whose
statistics may not be representative of a full season.

Variables that were not used as predictors, including the quarterback's
name, season, and team, were removed before model development. The
remaining dataset consisted of nine predictor variables: passing yards,
passing touchdowns, interceptions, completions, pass attempts, passing
EPA, rushing yards, rushing touchdowns, and completion percentage. The
response variable was the team's total number of regular-season wins.

Before fitting the predictive models, the predictor variables were
standardized so that each feature had a mean of zero and a standard
deviation of one. Standardization places variables measured on different
scales onto a common scale, improving model performance and ensuring
that variables with larger numerical ranges do not disproportionately
influence the results, particularly for regularization methods such as
Ridge and LASSO regression.

Finally, the cleaned dataset was randomly partitioned into training and
testing sets using an 90/10 split. The training set, containing 90% of
the observations, was used to fit and tune the predictive models, while
the remaining 10% served as an independent test set for evaluating model
performance on unseen data.

```{r}
# Removal of non-numeric variables.

drop_vars <- c("player_display_name", "season", "team")  
qb_season <- qb_season
qb_season <- qb_season[, !(names(qb_season) %in% drop_vars)]

# Scaling of numeric variables.

qb_scaled <- qb_season
predictor_cols <- setdiff(names(qb_scaled), "wins")

qb_scaled[predictor_cols] <- scale(qb_scaled[predictor_cols])

# Creation of train/test split.

set.seed(123)
train_index <- sample(seq_len(nrow(qb_scaled)), 0.8 * nrow(qb_scaled))

train <- qb_scaled[train_index, ]
test  <- qb_scaled[-train_index, ]

# Removal of missing values.

train <- na.omit(train)
test  <- na.omit(test)

# Simple regression model creation.

LinReg <- lm(wins ~ ., data = train)

# Matrix design in preparation for glmnet.

x_train <- model.matrix(wins ~ ., data = train)[, -1]
y_train <- train$wins
x_test_raw <- model.matrix(wins ~ ., data = test)[, -1]
x_test <- x_test_raw[, colnames(x_train), drop = FALSE]

# Ridge and Lasso creation including 10-fold cross-validation.

library(glmnet)
set.seed(123)
ridge_cv <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
set.seed(123)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)

# Prediction utilizing the test set.

ols_pred   <- predict(LinReg, newdata = test)
ridge_pred <- as.vector(predict(ridge_cv, newx = x_test, s = "lambda.min"))
lasso_pred <- as.vector(predict(lasso_cv, newx = x_test, s = "lambda.min"))

# Calculation of metrics.

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
mae  <- function(y, yhat) mean(abs(y - yhat))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# Evaluation based on sought-after metrics.

lin_coef <- coef(LinReg)

comparison <- data.frame(
  Variable = names(lin_coef),
  Linear = as.numeric(lin_coef),
  Ridge = as.numeric(coef(ridge_cv, s = "lambda.min")),
  LASSO = as.numeric(coef(lasso_cv, s = "lambda.min"))
)
kable(comparison, caption = "Table 2: Comparison of Linear, Ridge, and Lasso Regression Coefficients for Predicting Team Wins")


```

Table 2 shows that the linear regression model assigns larger
coefficients to features with potential correlations, such as passing
yards and attempts, which may indicate the presence of multicollinearity
among predictors. The Ridge regression model reduces the magnitude of
all coefficients while retaining all predictors in the model, helping to
reduce the impact of multicollinearity. The Lasso regression model
performs feature selection by shrinking the coefficients of passing
yards, interceptions, attempts, and rushing yards to zero. This suggests
that these variables provide less unique predictive information compared
to the other quarterback statistics included in the model.

An unexpected result from the linear regression model was the negative
coefficient for passing yards (-1.489). Although higher passing yard
totals are generally expected to contribute positively to team success,
this coefficient represents the relationship between passing yards and
wins while holding the other quarterback statistics constant. Since
passing yards are highly correlated with variables such as attempts and
completions, multicollinearity may have influenced the coefficient
estimate and resulted in a counterintuitive direction. The Ridge and
Lasso models reduce this issue by shrinking coefficients and reducing
the influence of correlated predictors.

```{r}
results <- data.frame(Model = c("OLS", "Ridge", "LASSO")
,
RMSE  = c(rmse(test$wins, ols_pred), rmse(test$wins, ridge_pred), rmse(test$wins, lasso_pred))
,
MAE = c(mae(test$wins, ols_pred), mae(test$wins, ridge_pred), mae(test$wins, lasso_pred))
,
R2 = c(r2(test$wins, ols_pred), r2(test$wins, ridge_pred), r2(test$wins, lasso_pred)))

kable(results, caption = "Table 3: Comparison of Model Performance Metrics for OLS, Ridge, and LASSO Regression" )
```

Table 3 compares the performance of the Ordinary Least Squares (OLS),
Ridge, and LASSO regression models using Root Mean Squared Error (RMSE),
Mean Absolute Error (MAE), and R². The OLS model achieved the best
overall performance with the lowest RMSE (2.072), lowest MAE (1.663),
and highest R² (0.589). Ridge and LASSO regression produced slightly
higher prediction errors and slightly lower explained variance. Although
regularization did not improve predictive performance in this dataset,
Ridge and LASSO helped reduce the impact of multicollinearity by
shrinking coefficient estimates and, in the case of LASSO, performing
feature selection.

```{r}
predictions <- data.frame(
  Actual = test$wins,
  OLS = ols_pred,
  Ridge = ridge_pred,
  LASSO = lasso_pred
)

head(predictions)

library(tidyr)
library(ggplot2)

predictions_long <- predictions %>%
  pivot_longer(
    cols = c(OLS, Ridge, LASSO),
    names_to = "Model",
    values_to = "Predicted"
  )

ggplot(predictions_long, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Actual vs Predicted Team Wins",
    x = "Actual Wins",
    y = "Predicted Wins"
  ) +
  theme_minimal()
```

```{r}
par(mfrow = c(1,2), cex.main = 0.9)

plot(ridge_cv)
title("Figure 5: Ridge Cross-Validation Curve")

plot(lasso_cv)
title("Figure 6: LASSO Cross-Validation Curve")
```

Figures 5 and 6 show how cross-validation was used to identify the
optimal regularization parameter ($\lambda$) for both Ridge and LASSO
regression models. The cross-validation curves demonstrate that
prediction error decreased as regularization was introduced, reached a
minimum at the selected $\lambda$, and increased with stronger
regularization. This pattern indicates that a considerable level of
regularization provided the best balance between model complexity and
predictive accuracy. The $\lambda$ values corresponding to the minimum
cross-validation error were used to fit the final Ridge and LASSO
models.

## Conclusion

This research evaluated LASSO and Ridge regression against a baseline
ordinary least squares model to predict team wins using quarterback
performance statistics. Prior to researching the linear regression model
compared to LASSO and Ridge regularized regression model, the
correlation plot from Figure 1 shows an almost perfect correlation
between passing statistics such as completions and pass attempts
(r=0.99) and passing yards (r=0.96) and EPA gave the strongest
correlation with team wins (r=0.66) and completion percentage gave a
moderately positive correlation with team wins (r=0.41). Figures 2 and 3
indicated possible strong positive correlation between team wins versus
passing yards and team wins versus passing touchdowns preemptively
suggesting these to be key variables in predicting team wins. However,
as shown in Table 2, the baseline Ordinary Least Squares model in this
research gave a negative coefficient for passing yards possibly being
influenced by multicollinearity due to passing yards being highly
correlated with other predictors such as completions and attempts. Ridge
regression reduced the value of all the coefficients without removing
predictors while LASSO regression shrunk the coefficients of passing
yards, interceptions, attempts and rushing yards to zero implicating
these variables to be the least important in predicting wins
comparatively with other quarterback variables. Which previously passing
yards was assumed to be a key variable but LASSO results show the
contrary. Although passing yards through LASSO and OLS analysis went
against previous assumptions, passing touchdowns maintained a strong
positive relationship with team wins. The combination of the distorted
OLS coefficient and elimination in LASSO, shows that passing yards, pass
attempts and rushing yards proved to not be key indicators of team
success. When evaluating performance prediction accuracy of OLS, LASSO
and Ridge regression as shown in Table 3, OLS achieved the highest
overall fit with the lowest RMSE and MAE and highest $R^2$ with Ridge
following second and LASSO following in third with the lowest fit out of
the other 2 models. Although OLS appears superior in this comparison
result, its previous erratic coefficient value from earlier diagnostics
causes concern for interpretability. Overall even though OLS seemed to
outperform LASSO and Ridge slightly in model performance metrics (RSME,
MAE and $R^2$), LASSO and Ridge with their shrinking techniques helped
to reduce the overall impact of multicollinearity more efficiently than
OLS suggesting that traditional linear regression methods could be
uncertain when facing complex and highly correlated variables. This
research helped to highlight the key variables of quarterback statistics
that contribute to team successes through regularized regression which
can prove beneficial for future predictions by providing an
interpretable foundation of how quarterback's performance impacts team
wins. 

## References
