---
title: "Sports Analytics Using Ridge/Lasso Regression Summer 2026"
subtitle: "Predicting Team Wins Based On QB Play"
author: "Jonathan Sears, Jasmine Cole, Noah Miller"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!
:::

## Introduction

Under the stadium lights, the smell of popcorn and hot dogs and the band
playing the “Fight song” as one final play runs the last 3 seconds of
the 4th quarter, that is the aesthetic of an American Football game.
Fans come together to support, cheer and hope for their team's victory.
But what if fans or people in general had the ability to go into a match
where they could predict the outcome? Being able to predict outcomes in
the National Football League (NFL) requires many different features and
data such as home field advantage, player performance, team statistics,
and game conditions which can cause “multicollinearity”.
Multicollinearity is the negative impact on the least square estimator
in regression due to correlation between variables [@mcdonald2009ridge]. 

With the implementation of a machine learning regression model such as
Lasso  (Least Absolute Shrinkage and Selection Operator), and Ridge we
can simplify the variables and focus only on the important aspects that
help to accurately predict wins. Lasso uses a shrinkage technique called
L1 regularization where it zeros out the least important variables,
leaving only variables that are optimal for accurate prediction. Ridge
regression uses a shrinkage technique called L2 regularization where it
helps reduce multicollinearity without removing variables from the
original set of independent variables. This helps with some of the
distortion of the predictors. 

Previous research examining LASSO regression has consistently
demonstrated its usefulness in datasets containing large numbers of
predictors and correlated variables. Across applications in sports
analytics, genetics, and other predictive modeling fields, LASSO has
been shown to improve model interpretability by shrinking insignificant
coefficients toward zero while retaining the most influential predictors
[@bhattacharyya2025lasso; @jana2021football; @lockhart2014lasso].
Such reduces model complexity and can improve predictive performance
compared to traditional regression methods [@alshqaq2023diagnostic; @bhattacharyya2025lasso; @lockhart2014lasso]. Studies have also shown
that regularization techniques such as Ridge regression can effectively
address multicollinearity by reducing coefficient variance without
removing predictors from the model. One limitation discussed within the
literature is that highly correlated variables can influence LASSO's
variable selection process, potentially causing relevant predictors to
be excluded [@lockhart2014lasso]. Overall, the findings suggest that
regularized regression methods provide a useful framework for prediction
problems involving numerous and potentially correlated variables, making
them applicable to NFL team performance analysis.

The main goal of this report is to examine the impact of
multicollinearity on predicting NFL team success and to examine whether
regularized regression models are better than a linear regression model.
This will be seen by comparing a linear regression, Ridge regression,
and Lasso regression ability to predict the number of regular-season
wins by an NFL team using that team’s quarterback's performance
statistics. Some of the predictor variables are the quarterback’s season
passing yards, touchdowns, interceptions and passing expected points 
added (EPA), a metric that measures how many net points a quarterback
adds to their team's scoring potential on passing plays, compared to the
league-average expectation for any given down, distance, and field
position. This report will also aim to find which quarterback
performance statistics are most influential in predicting team
performance. Understanding the impact of a quarterback’s performance on
a team's success and the most influential quarterback performance
statistics provide valuable information for NFL analysts, coaches, and
teams trying to find players that contribute to team success and improve
team performance.

The introduction should:

- Develop a storyline that captures attention and maintains interest.

- Your audience is your peers

- Clearly state the problem or question you're addressing.

<!-- -->

- Introduce why it is relevant needs.

- Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*. Topology can be used in machine
learning [@adams2021topology]

For Symbolic Regression [@wang2019symbolic] *This is my work and I want
to add more work...*

Cite new paper [@su2012linear]

Classification of Knee OA was studied here [@guida2021knee]

## Methods

- Detail the models or algorithms used.

- Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}$$ $W_n(x)$ *is the sum
of weights that belongs to all real numbers. Weights are positive
numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i$$

## Analysis and Results

### Data Exploration and Visualization

- Describe your data sources and collection process.

- Present initial findings and insights through visualizations.

- Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(corrplot)
library(nflreadr)
library(dplyr)
```

```{python}
import pandas as pd
```

```{r, warning=FALSE, message=FALSE, echo=TRUE}
# Load Data

years = 2015:2024
weekly = load_player_stats(years)
qbs <- weekly %>%
  filter(position == "QB")
qb_season <- qbs %>%
  group_by(player_display_name, season, team) %>%
  summarize(
    passing_yards = sum(passing_yards, na.rm = TRUE),
    passing_tds = sum(passing_tds, na.rm = TRUE),
    interceptions = sum(passing_interceptions, na.rm = TRUE),
    completions = sum(completions, na.rm = TRUE),
    attempts = sum(attempts, na.rm = TRUE),
    passing_epa = sum(passing_epa, na.rm = TRUE),
    rushing_yards = sum(rushing_yards, na.rm = TRUE),
    rushing_tds = sum(rushing_tds, na.rm = TRUE)
  )

qb_season <- qb_season %>%
  mutate(
    completion_pct = (completions / attempts) * 100
  )

write.csv(qb_season, "qb_season_stats.csv", row.names = FALSE)

schedules <- load_schedules(2015:2024)
team_wins <- schedules %>%
  filter(game_type == "REG") %>%
  mutate(
    home_win = ifelse(home_score > away_score, 1, 0),
    away_win = ifelse(away_score > home_score, 1, 0)
  )

home_wins <- team_wins %>%
  group_by(season, team = home_team) %>%
  summarize(wins = sum(home_win, na.rm = TRUE))

away_wins <- team_wins %>%
  group_by(season, team = away_team) %>%
  summarize(wins = sum(away_win, na.rm = TRUE))

team_total_wins <- bind_rows(home_wins, away_wins) %>%
  group_by(season, team) %>%
  summarize(wins = sum(wins))

qb_season <- qb_season %>%
  left_join(team_total_wins,
            by = c("season", "team"))

write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season <- qb_season %>%
  filter(attempts >= 200)


write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season1 <- qb_season %>%
  ungroup() %>%
  select(-player_display_name, -season, -team)
```

```{r}
correlation_matrix = round(cor(qb_season1, use = "complete.obs"), 2)
corrplot(correlation_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```

Figure 1: Correlation Plot between Variables

According to the correlation plot above, the passing EPA has the
strongest correlation with team wins (r = 0.66). Completion percentage
showed a moderate correlation with wins (r=0.41). The rushing statistics
like rushing touchdowns and rushing yards appeared to have the lowest
positive relationship with wins (average r= 0.22) while interceptions
had almost no linear relationship with wins (r=0.06). 

The analysis also identified substantial multicollinearity among several
passing statistics. Completions and attempts were nearly perfectly
correlated (r = 0.99), while passing yards showed very strong
correlations with completions (r = 0.97) and attempts (r = 0.96). These
findings suggest that some passing variables may provide redundant
information and should be carefully considered during model
development. 

### Modeling and Results

- Explain your data preprocessing and cleaning steps.

- Present your key findings in a clear and concise manner.

- Use visuals to support your claims.

- **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

- Summarize your key findings.

- Discuss the implications of your results.

## References

Gary C. McDonald. (2009). Ridge regression. WIREs Computational
Statistics, 1(1), 93–100. <https://doi.org/10.1002/wics.14>

Robert Tibshirani (1996). Regression shrinkage and selection via the
lasso. Journal of the Royal Statistical Society: Series B
(Methodological), 58(1), 267–288.
<https://doi.org/10.1111/j.2517-6161.1996.tb02080.x>

Gweon, J. (2024). Comparing different machine learning methods on NFL
defensive player performance prediction \[Master's report, Kansas State
University\]. K-State Research Exchange.[
https://krex.k-state.edu/items/b5cae4fb-f772-42ce-9dad-d7a84094e8bf](https://krex.k-state.edu/items/b5cae4fb-f772-42ce-9dad-d7a84094e8bf) 

Juuri, S. (2023). Predicting the results of NFL games using machine
learning \[Master's thesis, Aalto University\]. Aaltodoc.[
https://aaltodoc.aalto.fi/items/8a329777-bdfe-4800-8d8d-d6dc3e75be70](https://aaltodoc.aalto.fi/items/8a329777-bdfe-4800-8d8d-d6dc3e75be70) 

South, C., & Egros, E. (2020). Forecasting college football game
outcomes using modern modeling techniques. Journal of Sports Analytics,
6(1), 25-33.

Venkat Sai Tarun G, Sriramya P. Analyzing ola data for precise price
prediction using xgboost technique comparing with lasso regression. In:
Advances in Parallel Computing Algorithms, Tools and Paradigms. SAGE
Publications; 2026:360-366.
doi:[10.3233/APC220050](https://doi.org/10.3233/APC220050)

Laufer, B., Docherty, P. D., Murray, R., Krueger-Ziolek, S., Jalal, N.
A., Hoeflinger, F., Rupitsch, S. J., Reindl, L., & Moeller, K. (2023).
Sensor Selection for Tidal Volume Determination via Linear
Regression—Impact of Lasso versus Ridge Regression. Sensors (Basel,
Switzerland), 23(17), 7407. <https://doi.org/10.3390/s23177407>

Geraldo-Campos, L. A., Soria, J. J., & Pando-Ezcurra, T. (2022). Machine
learning for credit risk in the Reactive Peru Program: A comparison of
the Lasso and Ridge Regression models. Economies, 10(8), 1–21.
<https://doi.org/10.3390/economies10080188>

Alshqaq, S. S., & Abuzaid, A. H. (2023). An efficient method for
variable selection based on diagnostic-Lasso regression. Symmetry,
15(12), 2155.[
https://doi.org/10.3390/sym15122155](https://doi.org/10.3390/sym15122155)

Jana, A., & Hemalatha, S. (2021). Football player performance analysis
using particle swarm optimization and player value calculation using
regression. Journal of Physics: Conference Series, 1911(1), 012011.
<https://doi.org/10.1088/1742-6596/1911/1/012011>

Lockhart, R., Taylor, J., Tibshirani, R. J., & Tibshirani, R. (2014). A
significance test for the lasso. The Annals of Statistics, 42(2),
413–468. <https://doi.org/10.1214/13-AOS1175>

Bhattacharyya, J. (2025, January 23). LASSO regression – A procedural
improvement (SSRN Working Paper No. 5108591). SSRN.
https://doi.org/10.2139/ssrn.5108591

\

\
