---
title: "Sports Analytics Using Ridge/Lasso Regression Summer 2026"
subtitle: "Predicting Team Wins Based On QB Play"
author: "Jonathan Sears, Jasmine Cole, Noah Miller"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd` to edit)

::: callout-important
**Remember:** Your goal is to make your audience understand and care about your findings. By crafting a compelling story, you can effectively communicate the value of your data science project.

Carefully read this template since it has instructions and tips to writing!
:::

## Introduction

Under the stadium lights, the smell of popcorn and hot dogs and the band playing the “Fight song” as one final play runs the last 3 seconds of the 4th quarter, that is the aesthetic of an American Football game. Fans come together to support, cheer and hope for their team's victory. But what if fans or people in general had the ability to go into a match where they could predict the outcome? Being able to predict outcomes in the National Football League (NFL) requires many different features and data such as home field advantage, player performance, team statistics, and game conditions which can cause “multicollinearity”. Multicollinearity is the negative impact on the least square estimator in regression due to correlation between variables [@mcdonald2009ridge]. 

With the implementation of a machine learning regression model such as Lasso  (Least Absolute Shrinkage and Selection Operator), and Ridge we can simplify the variables and focus only on the important aspects that help to accurately predict wins. Lasso uses a shrinkage technique called L1 regularization where it zeros out the least important variables, leaving only variables that are optimal for accurate prediction. Ridge regression uses a shrinkage technique called L2 regularization where it helps reduce multicollinearity without removing variables from the original set of independent variables. This helps with some of the distortion of the predictors. 

Previous research examining LASSO regression has consistently demonstrated its usefulness in datasets containing large numbers of predictors and correlated variables. Across applications in sports analytics, genetics, and other predictive modeling fields, LASSO has been shown to improve model interpretability by shrinking insignificant coefficients toward zero while retaining the most influential predictors [@bhattacharyya2025lasso; @jana2021football; @lockhart2014lasso]. Such reduces model complexity and can improve predictive performance compared to traditional regression methods [@alshqaq2023diagnostic; @bhattacharyya2025lasso; @lockhart2014lasso]. Studies have also shown that regularization techniques such as Ridge regression can effectively address multicollinearity by reducing coefficient variance without removing predictors from the model. One limitation discussed within the literature is that highly correlated variables can influence LASSO's variable selection process, potentially causing relevant predictors to be excluded [@lockhart2014lasso]. Overall, the findings suggest that regularized regression methods provide a useful framework for prediction problems involving numerous and potentially correlated variables, making them applicable to NFL team performance analysis.

The main goal of this report is to examine the impact of multicollinearity on predicting NFL team success and to examine whether regularized regression models are better than a linear regression model. This will be seen by comparing a linear regression, Ridge regression, and Lasso regression ability to predict the number of regular-season wins by an NFL team using that team’s quarterback's performance statistics. Some of the predictor variables are the quarterback’s season passing yards, touchdowns, interceptions and passing expected points  added (EPA), a metric that measures how many net points a quarterback adds to their team's scoring potential on passing plays, compared to the league-average expectation for any given down, distance, and field position. This report will also aim to find which quarterback performance statistics are most influential in predicting team performance. Understanding the impact of a quarterback’s performance on a team's success and the most influential quarterback performance statistics provide valuable information for NFL analysts, coaches, and teams trying to find players that contribute to team success and improve team performance.

## Methods

Will update the methods this week. 7/7 JS

Across various predictive modeling applications ranging to sports outcomes in (South & Egros, 2020), sensor organization to ensure optimization (Laufer et al., 2023) and credit risks awareness in (Geraldo-Campos et al., 2022) LASSO, least absolute shrinkage and selection operator and ridge regression are both linear regression technique and methodologies that are designed to handle multicollinearity and mult-dimensional data. Both of these methodologies use a shrinkage tool that penalizes the regression coefficients. Both methods solve a modified optimization problem of the form:

\

where y is the response vector, X is the matrix of predictors, is the vector of the coefficients, P()is the penalty term, and 0 is the regularization parameter controlling the strength of the penalty. 

However, the difference between LASSO and Ridge regression comes from how their penalties are calculated mathematically. LASSO uses a L1 penalty norm, which shrinks the least important variables in the dataset to zero which allows for the selection of key effective variables 

(P()=j-1pj). Ridge regression uses a L2 penalty norm, which shrinks parameters close to zero reducing the variance (P()= j=1pj2) but where ridge differs from LASSO is that in the end and final model, ridge retains all variables . LASSO is a “step-up” from ridge allowing for key variable selection, where ridge can struggle with its limitations due to retention of all variables. This highlights that LASSO’s ability to produce a subset of “key-variables” from a high-dimensional dataset, makes it highly efficient. 

For example, sports outcomes in (South & Egros, 2020), LASSO was able to identify home-field advantage as a key-variable by narrowing down recruitment and box-score metrics. Similarly in (Laufer et al., 2023) LASSO was able to significantly reduce computational complexity that identified an optimal configuration of motion sensors. However it is important to note that LASSO has experienced some limitations in its performance ability in variable selection specifically with highly correlated variables and complex non-linear interactions. For example in (Tarun & Sriramya, 2026), this study used lasso regression against XGBoost to determine which methodology proved to be the most accurate at predicting fare prices for online cab services which XGBoost outperformed LASSO because “LASSO failed to make the grouped selection as it only took one variable from a group and ignored the remaining variables”(Tarun & Sriramya, 2026). However, despite this specific limitation, across multiple studies LASSO proves to be consistent in its ability to minimize prediction error and utilize mathematical frameworks to isolate key critical variables.

For both regression methods, the regularization parameter () was selected using a 10-fold cross validation, ensuring a practical 90/10 train-test split. Within the function, the dataset is divided into ten equally sized subsets, where nine folds are used to train the model while the remaining fold is reserved for validation (testing). The process is repeated until each fold has served as the validation set exactly once, and the resulting validation errors are averaged to estimate the model's predictive performance. By evaluating each candidate value of the regularization parameter () across all ten folds, what is identified is the value that minimizes the cross-validated error. The use of 10-fold cross-validation provides a balance between computational efficiency and reliable model evaluation, allowing nearly all observations to contribute to both model training and validation throughout the procedure, without heightened risk of overfitting or underfitting.

To evaluate the predictive performance of the Lasso and Ridge regression models, RMSE, MAE, and R² were calculated using the cross-validated predictions. RMSE and MAE quantify the average prediction error, with RMSE placing greater emphasis on larger errors due to the squared error term, while MAE provides a measure that is less sensitive to extreme observations. R², however, measures the proportion of variation in the response variable explained by the model. Lower RMSE and MAE values, coupled with a higher R², indicate a model that more effectively balances prediction accuracy with the coefficient shrinkage introduced through regularization.

The R programming language was the primary software used in this analysis. Data for the analysis was sourced from the nflreadr package; a repository of NFL data easily accessible within the R environment. For data preprocessing, the Python programming language package pandas was used within the R environment, as well as the native dplyr package. The corrplot, car, and ggplot2 packages were used for the preliminary analysis and analysis of the final models. Computation and selection of the LASSO and Ridge-based models was completed through use of glmnet.

\

- Detail the models or algorithms used.

- Justify your choices based on the problem and data.

*The common non-parametric regression model is* $Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is unknown and* $\varepsilon_i$ *some errors. With the help of this definition, we can create the estimation for local averaging i.e.* $m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$ *is near to* $x$*. In other words, this means that we are discovering the line through the data points with the help of surrounding data points. The estimation formula is printed below [@R-base]:*

$$M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}$$ $W_n(x)$ *is the sum of weights that belongs to all real numbers. Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i$$

## Analysis and Results

### Data Exploration and Visualization

A study was conducted to determine how quarterback (QB) season statistics correlates with how many wins a team has in that season. To determine how QB statistics correlates with team wins in a season, the dataset was created using publicly available NFL data accessed through the R package [nflreadr]{.underline}. Weekly quarterback player statistics and NFL game schedules were collected for the 2015 through 2024 regular seasons.

Quarterback performance data were obtained using the `load_player_stats()` function. The dataset was filtered to include only players whose position was quarterback (QB). Weekly statistics were then aggregated into season-level totals for each quarterback and team. The aggregated variables included passing yards, passing touchdowns, interceptions, completions, pass attempts, passing EPA (Expected Points Added), rushing yards, and rushing touchdowns. In addition, completion percentage was calculated by dividing total completions by total pass attempts and multiplying by 100.

Team performance data were collected using the `load_schedules()` function. Only regular-season games were included by filtering for `game_type == "REG"`. Wins were calculated separately for home and away teams by comparing the final scores of each game. Home and away wins were then combined to compute each team's total number of regular-season victories for every season from 2015 through 2024.

The team win totals were merged with the quarterback season statistics using season and team as matching variables. Finally, quarterbacks with fewer than 200 pass attempts in a season were removed to ensure that the analysis included only quarterbacks with substantial playing time, reducing the influence of backup quarterbacks and small sample sizes.

The final dataset contains one observation for each qualifying quarterback-season and includes both individual performance statistics and the corresponding team's regular-season win total. The description of variables are listed below in the chart.

```{r}
library(dplyr)
library(knitr)

variable_table <- data.frame(
  Variable = c(
    "passing_yards",
    "passing_tds",
    "interceptions",
    "completions",
    "attempts",
    "passing_epa",
    "rushing_yards",
    "rushing_tds",
    "completion_pct",
    "wins"
  ),
  Description = c(
    "Total number of passing yards recorded by the quarterback during the season.",
    "Total number of passing touchdowns thrown by the quarterback during the season.",
    "Total number of interceptions thrown by the quarterback during the season.",
    "Total number of completed passes by the quarterback during the season.",
    "Total number of pass attempts by the quarterback during the season.",
    "Total Expected Points Added (EPA) generated through passing plays during the season.",
    "Total number of rushing yards gained by the quarterback during the season.",
    "Total number of rushing touchdowns scored by the quarterback during the season.",
    "Completion percentage, calculated as (completions ÷ attempts) × 100.",
    "Total number of regular-season wins recorded by the quarterback's team (response variable)."
  ),
  Type = c(
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Predictor",
    "Response"
  )
)

kable(variable_table,
      caption = "Description of Variables Used in the Analysis")
```

- Describe your data sources and collection process.

- Present initial findings and insights through visualizations.

- Highlight unexpected patterns or anomalies.

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(corrplot)
library(nflreadr)
library(dplyr)

```

```{python}
import pandas as pd
```

```{r, warning=FALSE, message=FALSE, echo=TRUE}
# Load Data

years = 2015:2024
weekly = load_player_stats(years)
qbs <- weekly %>%
  filter(position == "QB")
qb_season <- qbs %>%
  group_by(player_display_name, season, team) %>%
  summarize(
    passing_yards = sum(passing_yards, na.rm = TRUE),
    passing_tds = sum(passing_tds, na.rm = TRUE),
    interceptions = sum(passing_interceptions, na.rm = TRUE),
    completions = sum(completions, na.rm = TRUE),
    attempts = sum(attempts, na.rm = TRUE),
    passing_epa = sum(passing_epa, na.rm = TRUE),
    rushing_yards = sum(rushing_yards, na.rm = TRUE),
    rushing_tds = sum(rushing_tds, na.rm = TRUE)
  )

qb_season <- qb_season %>%
  mutate(
    completion_pct = (completions / attempts) * 100
  )

write.csv(qb_season, "qb_season_stats.csv", row.names = FALSE)

schedules <- load_schedules(2015:2024)
team_wins <- schedules %>%
  filter(game_type == "REG") %>%
  mutate(
    home_win = ifelse(home_score > away_score, 1, 0),
    away_win = ifelse(away_score > home_score, 1, 0)
  )

home_wins <- team_wins %>%
  group_by(season, team = home_team) %>%
  summarize(wins = sum(home_win, na.rm = TRUE))

away_wins <- team_wins %>%
  group_by(season, team = away_team) %>%
  summarize(wins = sum(away_win, na.rm = TRUE))

team_total_wins <- bind_rows(home_wins, away_wins) %>%
  group_by(season, team) %>%
  summarize(wins = sum(wins))

qb_season <- qb_season %>%
  left_join(team_total_wins,
            by = c("season", "team"))

write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season <- qb_season %>%
  filter(attempts >= 200)


write.csv(qb_season,
          "qb_season_dataset.csv",
          row.names = FALSE)

qb_season1 <- qb_season %>%
  ungroup() %>%
  select(-player_display_name, -season, -team)
```

```{r}
correlation_matrix = round(cor(qb_season1, use = "complete.obs"), 2)
corrplot(correlation_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```

Figure 1: Correlation Plot between Variables

According to the correlation plot above, the passing EPA has the strongest correlation with team wins (r = 0.66). Completion percentage showed a moderate correlation with wins (r=0.41). The rushing statistics like rushing touchdowns and rushing yards appeared to have the lowest positive relationship with wins (average r= 0.22) while interceptions had almost no linear relationship with wins (r=0.06). 

The analysis also identified substantial multicollinearity among several passing statistics. Completions and attempts were nearly perfectly correlated (r = 0.99), while passing yards showed very strong correlations with completions (r = 0.97) and attempts (r = 0.96). These findings suggest that some passing variables may provide redundant information and should be carefully considered during model development. 

```{r}
library(ggplot2)

ggplot(qb_season1, aes(x = passing_tds, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm",color = "red") + labs(x="Passing touchdowns", y="Team wins", title = "Passing touchdowns vs team wins")
```

```{r}
ggplot(qb_season1, aes(x = passing_yards, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm") + labs(x="Passing yards", y="Team wins", title = "Passing yards vs team wins")
```

Both scatterplots show that passing yards and passing touchdowns have a relative strong positive correlation to team wins. These scatterplots show that both variables, passing yards and passing touchdowns have a positive effect on team wins. The passing touchdowns versus team wins scatterplot shows a stronger relationship with less variability and spread in the data comparatively to passing yards versus team wins. The conclusion can be made that as passing touchdowns increase, team wins also increase and similarly as passing yards increase, team wins increase. This conclusion is beneficial in highlightly possible key variables that strongly influence team wins. \

```{r}
VIF_Visualization_Model <- lm(wins ~ passing_yards + passing_tds + interceptions + completions + attempts + passing_epa + rushing_yards + rushing_tds + completion_pct, data = qb_season1)
library(car)
VIF_Visualization_Values <- vif(VIF_Visualization_Model)
VIF_Visualization_Values
VIF_Visualization_Dataframe <- data.frame(Variable = names(VIF_Visualization_Values), VIF = as.numeric(VIF_Visualization_Values))
library(ggplot2)
ggplot(VIF_Visualization_Dataframe, aes(x = reorder(Variable, VIF), y = VIF)) + geom_col(fill = "steelblue") + geom_hline(yintercept = 5, linetype = "dashed", color = "red") + coord_flip() + labs(title = "Variance Inflation Factors (VIFs) of Dataset Predictors", x = "Predictor", y = "VIF") + theme_minimal()

```

Variance Inflation Factors (VIFs) of the predictors against the response variable (wins) are modeled based on a simple linear regression created for the purposes of this visualization. To combat inflated variances due to predictor multicollinearity, a VIF threshold of 5 was used, exhibited by the red-dashed line running along the y-axis. Predictors such as interceptions, rushing_tds, and rushing_yards satisfy this threshold. Although completion_pct, passing_tds, and passing_epa do not meet the threshold, they are close enough to at least consider in a final model if other evidence supports their inclusion. Completions, attempts, and passing_yards, however, indicate such high variances that they should likely not be used when refining a model.

### Modeling and Results

After collecting the quarterback and team data, several preprocessing steps were performed to prepare the dataset for analysis.

First, weekly quarterback statistics were aggregated into season-level totals for each quarterback and team from the 2015–2024 NFL regular seasons. Team regular-season win totals were calculated separately from the schedule data by summing home and away victories and then merged with the quarterback statistics using the season and team as matching variables.

A new variable, **completion percentage**, was created by dividing total completions by total pass attempts and multiplying by 100. This provided an additional measure of passing efficiency beyond the raw completion and attempt totals.

To ensure that only quarterbacks with meaningful playing time were included, seasons in which a quarterback attempted fewer than 200 passes were removed from the dataset. This filtering step reduced the influence of backup quarterbacks and players with limited appearances whose statistics may not be representative of a full season.

Variables that were not used as predictors, including the quarterback's name, season, and team, were removed before model development. The remaining dataset consisted of nine predictor variables: passing yards, passing touchdowns, interceptions, completions, pass attempts, passing EPA, rushing yards, rushing touchdowns, and completion percentage. The response variable was the team's total number of regular-season wins.

Before fitting the predictive models, the predictor variables were standardized so that each feature had a mean of zero and a standard deviation of one. Standardization places variables measured on different scales onto a common scale, improving model performance and ensuring that variables with larger numerical ranges do not disproportionately influence the results, particularly for regularization methods such as Ridge and LASSO regression.

Finally, the cleaned dataset was randomly partitioned into training and testing sets using an 80/20 split. The training set, containing 80% of the observations, was used to fit and tune the predictive models, while the remaining 20% served as an independent test set for evaluating model performance on unseen data.

```{r}
# Removal of non-numeric variables.

drop_vars <- c("player_display_name", "season", "team")  
qb_season <- qb_season
qb_season <- qb_season[, !(names(qb_season) %in% drop_vars)]

# Scaling of numeric variables.

qb_scaled <- qb_season
numeric_cols <- sapply(qb_scaled, is.numeric)
qb_scaled[numeric_cols] <- scale(qb_scaled[numeric_cols])

# Creation of train/test split.

set.seed(123)
train_index <- sample(seq_len(nrow(qb_scaled)), 0.8 * nrow(qb_scaled))

train <- qb_scaled[train_index, ]
test  <- qb_scaled[-train_index, ]

# Removal of missing values.

train <- na.omit(train)
test  <- na.omit(test)

# Simple regression model creation.

LinReg <- lm(wins ~ ., data = train)

# Matrix design in preparation for glmnet.

x_train <- model.matrix(wins ~ ., data = train)[, -1]
y_train <- train$wins
x_test_raw <- model.matrix(wins ~ ., data = test)[, -1]
x_test <- x_test_raw[, colnames(x_train), drop = FALSE]

# Ridge and Lasso creation including 10-fold cross-validation.

library(glmnet)
set.seed(123)
ridge_cv <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
set.seed(123)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)


lin_coef <- coef(LinReg)

comparison <- data.frame(
  Variable = names(lin_coef),
  Linear = as.numeric(lin_coef),
  Ridge = as.numeric(coef(ridge_cv, s = "lambda.min")),
  Lasso = as.numeric(coef(lasso_cv, s = "lambda.min"))
)

comparison

```

```{r}
# Prediction utilizing the test set.

ols_pred   <- predict(LinReg, newdata = test)
ridge_pred <- as.vector(predict(ridge_cv, newx = x_test, s = "lambda.min"))
lasso_pred <- as.vector(predict(lasso_cv, newx = x_test, s = "lambda.min"))

# Calculation of metrics.

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
mae  <- function(y, yhat) mean(abs(y - yhat))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# Evaluation based on sought-after metrics.

results <- data.frame(Model = c("OLS", "Ridge", "Lasso")
,
RMSE  = c(rmse(test$wins, ols_pred), rmse(test$wins, ridge_pred), rmse(test$wins, lasso_pred))
,
MAE = c(mae(test$wins, ols_pred), mae(test$wins, ridge_pred), mae(test$wins, lasso_pred))
,
R2 = c(r2(test$wins, ols_pred), r2(test$wins, ridge_pred), r2(test$wins, lasso_pred)))

results
```

### Explain your data preprocessing and cleaning steps.

- Present your key findings in a clear and concise manner.

- Use visuals to support your claims.

- **Tell a story about what the data reveals.**Conclusion

<!-- -->

- Summarize your key findings.

- Discuss the implications of your results.

## References

Gary C. McDonald. (2009). Ridge regression. WIREs Computational Statistics, 1(1), 93–100. <https://doi.org/10.1002/wics.14>

Robert Tibshirani (1996). Regression shrinkage and selection via the lasso. Journal of the Royal Statistical Society: Series B (Methodological), 58(1), 267–288. <https://doi.org/10.1111/j.2517-6161.1996.tb02080.x>

Gweon, J. (2024). Comparing different machine learning methods on NFL defensive player performance prediction \[Master's report, Kansas State University\]. K-State Research Exchange.<https://krex.k-state.edu/items/b5cae4fb-f772-42ce-9dad-d7a84094e8bf> 

Juuri, S. (2023). Predicting the results of NFL games using machine learning \[Master's thesis, Aalto University\]. Aaltodoc.<https://aaltodoc.aalto.fi/items/8a329777-bdfe-4800-8d8d-d6dc3e75be70> 

South, C., & Egros, E. (2020). Forecasting college football game outcomes using modern modeling techniques. Journal of Sports Analytics, 6(1), 25-33.

Venkat Sai Tarun G, Sriramya P. Analyzing ola data for precise price prediction using xgboost technique comparing with lasso regression. In: Advances in Parallel Computing Algorithms, Tools and Paradigms. SAGE Publications; 2026:360-366. doi:[10.3233/APC220050](https://doi.org/10.3233/APC220050)

Laufer, B., Docherty, P. D., Murray, R., Krueger-Ziolek, S., Jalal, N. A., Hoeflinger, F., Rupitsch, S. J., Reindl, L., & Moeller, K. (2023). Sensor Selection for Tidal Volume Determination via Linear Regression—Impact of Lasso versus Ridge Regression. Sensors (Basel, Switzerland), 23(17), 7407. <https://doi.org/10.3390/s23177407>

Geraldo-Campos, L. A., Soria, J. J., & Pando-Ezcurra, T. (2022). Machine learning for credit risk in the Reactive Peru Program: A comparison of the Lasso and Ridge Regression models. Economies, 10(8), 1–21. <https://doi.org/10.3390/economies10080188>

Alshqaq, S. S., & Abuzaid, A. H. (2023). An efficient method for variable selection based on diagnostic-Lasso regression. Symmetry, 15(12), 2155.<https://doi.org/10.3390/sym15122155>

Jana, A., & Hemalatha, S. (2021). Football player performance analysis using particle swarm optimization and player value calculation using regression. Journal of Physics: Conference Series, 1911(1), 012011. <https://doi.org/10.1088/1742-6596/1911/1/012011>

Lockhart, R., Taylor, J., Tibshirani, R. J., & Tibshirani, R. (2014). A significance test for the lasso. The Annals of Statistics, 42(2), 413–468. <https://doi.org/10.1214/13-AOS1175>

Bhattacharyya, J. (2025, January 23). LASSO regression – A procedural improvement (SSRN Working Paper No. 5108591). SSRN. https://doi.org/10.2139/ssrn.5108591

\

\
