---
title: "Predicting Team Wins Based on QB Play"
subtitle: "Using Ridge and LASSO Regression to Address Multicollinearity"
author: "Jonathan Sears, Jasmine Cole, Noah Miller (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction {.smaller}

- Develop a storyline that captures attention and maintains interest.

- Your audience is your peers

- Clearly state the problem or question you're addressing.

- Introduce why it is relevant needs.

- Provide an overview of your approach.

In kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]\*

## Methods

- Detail the models or algorithms used.

- Justify your choices based on the problem and data.

## Data Exploration and Visualization (Data Source and Collection Process)

- Dataset was created using R package **nflreadr**

- Collected quarterback (QB) performance data using
  `load_player_stats()` function for 2015-2024 seasons

- Aggregated weekly statistics into season stats

- Collected win totals using the `load_schedules()` function and counted
  number of wins for home and away games for each team.

- Joined QB statistics with team wins by matching season and team as
  matching variables

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how quarterback (QB) season
statistics correlates with how many wins a team has in that season.

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(corrplot)
library(nflreadr)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggplot2)
```

```{r, warning=FALSE, echo=F}
# Load Data
variable_table <- data.frame(
  Variable = c(
    "Passing Yards",
    "Passing TDs",
    "Interceptions",
    "Completions",
    "Attempts",
    "Passing EPA",
    "Rushing Yards",
    "Rushing TDs",
    "Completion %",
    "Wins"
  ),
  Type = c(
    rep("Predictor", 9),
    "Response"
  )
)

knitr::kable(
  variable_table
) %>%
  kableExtra::kable_styling(
    font_size = 20,
    full_width = FALSE
  )

```

## Data Exploration and Visualization {.smaller}

```{r}
years = 2015:2024
weekly = load_player_stats(years)
qbs <- weekly %>%
  filter(position == "QB")
qb_season <- qbs %>%
  group_by(player_display_name, season, team) %>%
  summarize(
    passing_yards = sum(passing_yards, na.rm = TRUE),
    passing_tds = sum(passing_tds, na.rm = TRUE),
    interceptions = sum(passing_interceptions, na.rm = TRUE),
    completions = sum(completions, na.rm = TRUE),
    attempts = sum(attempts, na.rm = TRUE),
    passing_epa = sum(passing_epa, na.rm = TRUE),
    rushing_yards = sum(rushing_yards, na.rm = TRUE),
    rushing_tds = sum(rushing_tds, na.rm = TRUE)
  )

qb_season <- qb_season %>%
  mutate(
    completion_pct = (completions / attempts) * 100
  )


schedules <- load_schedules(2015:2024)
team_wins <- schedules %>%
  filter(game_type == "REG") %>%
  mutate(
    home_win = ifelse(home_score > away_score, 1, 0),
    away_win = ifelse(away_score > home_score, 1, 0)
  )

home_wins <- team_wins %>%
  group_by(season, team = home_team) %>%
  summarize(wins = sum(home_win, na.rm = TRUE))

away_wins <- team_wins %>%
  group_by(season, team = away_team) %>%
  summarize(wins = sum(away_win, na.rm = TRUE))

team_total_wins <- bind_rows(home_wins, away_wins) %>%
  group_by(season, team) %>%
  summarize(wins = sum(wins))

qb_season <- qb_season %>%
  left_join(team_total_wins,
            by = c("season", "team"))



qb_season <- qb_season %>%
  filter(attempts >= 200)




qb_season1 <- qb_season %>%
  ungroup() %>%
  select(-player_display_name, -season, -team)

correlation_matrix = round(cor(qb_season1, use = "complete.obs"), 2)
corrplot(correlation_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45)
```

## Data Exploration and Visualization {.smaller}

::: {layout-ncol="2"}
```{r, echo=FALSE}
#| fig-width: 6
#| fig-height: 5

ggplot(qb_season1, aes(x = passing_tds, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm",color = "red") + labs(x="Passing touchdowns", y="Team wins", title = "Passing Touchdowns vs Team Wins")
```

```{r}
#| fig-width: 6
#| fig-height: 5

ggplot(qb_season1, aes(x = passing_yards, y = wins)) + geom_point(alpha = 0.7) + geom_smooth(method = "lm") + labs(x="Passing yards", y="Team wins", title = "Passing Yards vs Team Wins")
```
:::

## Data Exploration and Visualization {.smaller}

```{r}
VIF_Visualization_Model <- lm(wins ~ passing_yards + passing_tds + interceptions + completions + attempts + passing_epa + rushing_yards + rushing_tds + completion_pct, data = qb_season1)
library(car)
VIF_Visualization_Values <- vif(VIF_Visualization_Model)
VIF_Visualization_Values
VIF_Visualization_Dataframe <- data.frame(Variable = names(VIF_Visualization_Values), VIF = as.numeric(VIF_Visualization_Values))
library(ggplot2)
ggplot(VIF_Visualization_Dataframe, aes(x = reorder(Variable, VIF), y = VIF)) + geom_col(fill = "steelblue") + geom_hline(yintercept = 5, linetype = "dashed", color = "red") + coord_flip() + labs(title = "Variance Inflation Factors (VIFs) of Dataset Predictors", x = "Predictor", y = "VIF") + theme_minimal()

```

## Modeling and Results (Data Preprocessing)

- Created a new variable called completion percentage

- QBs with fewer than 200 pass attempts were removed

- All predictor variables were standardized

- The data set was was split into a 90% training set, 10% test set

## Modeling and Results

```{r}
# Removal of non-numeric variables.

drop_vars <- c("player_display_name", "season", "team")  
qb_season <- qb_season
qb_season <- qb_season[, !(names(qb_season) %in% drop_vars)]

# Scaling of numeric variables.

qb_scaled <- qb_season
predictor_cols <- setdiff(names(qb_scaled), "wins")

qb_scaled[predictor_cols] <- scale(qb_scaled[predictor_cols])

# Creation of train/test split.

set.seed(123)
train_index <- sample(seq_len(nrow(qb_scaled)), 0.9 * nrow(qb_scaled))

train <- qb_scaled[train_index, ]
test  <- qb_scaled[-train_index, ]

# Removal of missing values.

train <- na.omit(train)
test  <- na.omit(test)

# Simple regression model creation.

LinReg <- lm(wins ~ ., data = train)

# Matrix design in preparation for glmnet.

x_train <- model.matrix(wins ~ ., data = train)[, -1]
y_train <- train$wins
x_test_raw <- model.matrix(wins ~ ., data = test)[, -1]
x_test <- x_test_raw[, colnames(x_train), drop = FALSE]

# Ridge and Lasso creation including 10-fold cross-validation.

library(glmnet)
set.seed(123)
ridge_cv <- cv.glmnet(x_train, y_train, alpha = 0, nfolds = 10)
set.seed(123)
lasso_cv <- cv.glmnet(x_train, y_train, alpha = 1, nfolds = 10)


# Training predictions
ols_train_pred <- predict(LinReg, newdata = train)
ridge_train_pred <- as.vector(predict(ridge_cv, newx = x_train, s = "lambda.min"))
lasso_train_pred <- as.vector(predict(lasso_cv, newx = x_train, s = "lambda.min"))


# Prediction utilizing the test set.

ols_pred   <- predict(LinReg, newdata = test)
ridge_pred <- as.vector(predict(ridge_cv, newx = x_test, s = "lambda.min"))
lasso_pred <- as.vector(predict(lasso_cv, newx = x_test, s = "lambda.min"))

# Calculation of metrics.

rmse <- function(y, yhat) sqrt(mean((y - yhat)^2))
mae  <- function(y, yhat) mean(abs(y - yhat))
r2   <- function(y, yhat) 1 - sum((y - yhat)^2) / sum((y - mean(y))^2)

# Evaluation based on sought-after metrics.

lin_coef <- coef(LinReg)

comparison <- data.frame(
  Variable = names(lin_coef),
  Linear = as.numeric(lin_coef),
  Ridge = as.numeric(coef(ridge_cv, s = "lambda.min")),
  LASSO = as.numeric(coef(lasso_cv, s = "lambda.min"))
)

comparison <- comparison |>
  mutate(
    Linear = round(Linear, 3),
    Ridge = round(Ridge, 3),
    LASSO = round(LASSO, 3)
  )

knitr::kable(
  comparison,
  col.names = c("Variable", "OLS", "Ridge", "LASSO")
) %>%
  kable_styling(
    font_size = 20,
    position = "center",
    full_width = FALSE
  )
```

## Modeling and Results

```{r}
num_predictors <- ncol(x_train)

ols_predictors <- num_predictors
ridge_predictors <- num_predictors

# Count nonzero coefficients (excluding intercept)
lasso_predictors <- sum(coef(lasso_cv, s = "lambda.min")[-1] != 0)

results <- data.frame(
  Model = c("OLS", "Ridge", "LASSO"),

  Predictors = c(
    ols_predictors,
    ridge_predictors,
    lasso_predictors
  ),

  Train_RMSE = c(
    rmse(train$wins, ols_train_pred), 
    rmse(train$wins, ridge_train_pred), 
    rmse(train$wins, lasso_train_pred)
  ),

  Train_MAE = c(
    mae(train$wins, ols_train_pred),
    mae(train$wins, ridge_train_pred),
    mae(train$wins, lasso_train_pred)
  ),

  Train_R2 = c(
    r2(train$wins, ols_train_pred),
    r2(train$wins, ridge_train_pred),
    r2(train$wins, lasso_train_pred)
  ),

  Test_RMSE = c(
    rmse(test$wins, ols_pred), 
    rmse(test$wins, ridge_pred), 
    rmse(test$wins, lasso_pred)
  ),

  Test_MAE = c(
    mae(test$wins, ols_pred),
    mae(test$wins, ridge_pred),
    mae(test$wins, lasso_pred)
  ),

  Test_R2 = c(
    r2(test$wins, ols_pred),
    r2(test$wins, ridge_pred),
    r2(test$wins, lasso_pred)
  )
)

knitr::kable(
  results,
  digits = 3
) %>%
  kable_styling(
    font_size = 20,
    position = "center",
    full_width = FALSE
  )
```

## Modeling and Results

```{r}
predictions <- data.frame(
  Actual = test$wins,
  OLS = ols_pred,
  Ridge = ridge_pred,
  LASSO = lasso_pred
)

#head(predictions)

library(tidyr)
library(ggplot2)

predictions_long <- predictions %>%
  pivot_longer(
    cols = c(OLS, Ridge, LASSO),
    names_to = "Model",
    values_to = "Predicted"
  )

ggplot(predictions_long, aes(x = Actual, y = Predicted, color = Model)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed"
  ) +
  labs(
    title = "Actual vs Predicted Team Wins",
    x = "Actual Wins",
    y = "Predicted Wins"
  ) +
  theme_minimal()
```

## Modeling and Results

```{r}
par(mfrow = c(1,2), cex.main = 0.9)

plot(ridge_cv)
title("Ridge Cross-Validation Curve")

plot(lasso_cv)
title("LASSO Cross-Validation Curve")
```

## Conclusion

- Summarize your key findings.

- Discuss the implications of your results.

## References

:::
